### 2023-09-01 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@7deac7d13ef5a3958af7107cc311210c3f33b74b 🚀

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic
interactions.\textsuperscript{\textbf{Thompson2005GeoMos?}} The GMTC
posits that coevolutionary processes among
pairs\textsuperscript{\textbf{Thompson1994CoePro?}} or
complexes\textsuperscript{\textbf{Janzen1980WheIt?}} of species are
structured in space by the rippling effects of abiotic conditions onto
evolutionary mechanisms, giving rise to fragmented systems with
different ecologies over large spatial
extents.\textsuperscript{\textbf{Price2002MacThe?}} The GMTC predicts a
spatial fragmentation of coevolutionary dynamics under the joint action
of three processes:\textsuperscript{\textbf{Gomulkiewicz2007DosDon?}}
coevolutionary hot- and coldspots, which appear when the intensity of
\emph{interaction} (in terms of reciprocal fitness consequences) varies
spatially; selection mosaics, wherein the intensity of \emph{selection}
varies across space, driven by both the biotic complexity of the
community (locally diverse hosts and viruses are more biotically
complex) and the local favorability of the
environment;\textsuperscript{\textbf{Thrall2007CoeSym?}} and trait
remixing, which occurs when coevolutionary dynamics change when
community-level \emph{functional traits} change through meta-community
dynamics.

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. In their bat reservoirs, coronaviruses evolve through a mix
of host jumps, recombination among disparate lineages, and, to a lesser
degree, co-divergence with their
hosts---\textsuperscript{\textbf{Anthony2017GloPat?}}a mix of mechanisms
that creates a complex and nonlinear relationship between host diversity
and viral emergence. Working from a recently published database of bat
hosts of betacoronaviruses, we test whether spatial structure in
bat-betacoronavirus coevolution is identifiable at a global scale.
Aiming to explain these patterns, we develop a generalized framework for
applying the GMTC to host-virus interactions, with a specific emphasis
on the potential to create independent coevolutionary dynamics (and
therefore spatial fragmentation in risk) through heterogeneity. We
develop a trivariate risk assessment system that connects each GMTC
mechanism to a quantifiable aspect of host-virus interactions: (i) viral
sharing rates in host communities, representing the strength of
potential interaction between viruses and any one host (i.e., places
where viruses undergo constant host switching may be coevolutionary
coldspots); (ii) the phylogenetic diversity of hosts, as a proxy for
variation in the immunological mechanisms that antagonize viruses (i.e.,
the selection mosaic); and (iii) the local uniqueness of the bat
community, representing the potential for viruses to be exposed to novel
host traits (e.g., variation in receptor sequences). Together, we argue
that these can be used to identify and map the evolutionary drivers
that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?}}
Projecting bat and betacoronavirus phylogeny over space
(fig.~\ref{fig:biogeo}), we find support for the idea that bat community
assembly is directly responsible for a global mosaic of viral evolution.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia; (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and Latin America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of multiple
betacoronavirus subgenera, including nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{\textbf{Ruiz-Aravena2022EcoEvo?}}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{\textbf{Anthony2017GloPat?}}a property
that would theoretically allow for substantial broad divergence in their
biogeography. However, host jumps at the family level or higher are
relatively rare and significant events in coronavirus evolutionary
history;\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Latinne2020OriCro?}}
as a result, the mosaic of betacoronavirus phylogeography is assembled
from a set of overlapping smaller coevolutionary systems, superimposed
in space and filtered by the importance of different subgroups in local
host communities. For example, the most speciose and cosmopolitan family
of bats, the vesper bats (Vespertilionidae), are considered the primary
hosts of the subgenus \emph{Merbecovirus} (MERS-like
viruses);\textsuperscript{\textbf{Latinne2020OriCro?},\textbf{Ruiz-Aravena2022EcoEvo?}}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{\textbf{Anthony2013CorBat?},\textbf{Goes2013NovBat?},\textbf{Goes2016GenDiv?},\textbf{Brandao2008CorDet?}}
At the coarsest scale, these heterogeneities are lost, and
betacoronavirus biogeography tracks the deep rifts in bat evolutionary
history---but within broad regions, the component coevolutionary systems
may have very different dynamics.

@@

Bats---and the spillover of their viruses---are also sensitive to
anthropogenic factors others than climate change, including
deforestation and other kinds of habitat loss, increased stress, and
greater contact with potential bridge hosts like domesticated
species.\textsuperscript{\textbf{Alves2018GeoVar?},\textbf{Treitler2016EffLoc?},\textbf{Rulli2021LanCha?},\textbf{Mendenhall2014PreBio?}}
This represents a challenge for both conservation strategies and
pandemic prevention,\textsuperscript{\textbf{Amman2011InvRol?}} but
identifying areas at risk, and protecting the health of bats and
ecosystems within those zones, can be a win-win intervention for
both.\textsuperscript{\textbf{Hopkins2021HowIde?},\textbf{Plowright2021LanUse?},\textbf{OHHLEP2022OneHea?}}
As we scale these predictions down in space to finer spatial resolutions
to guide public health actions,\textsuperscript{5} the incorporation of
human activity predictors will become more
importyant.\textsuperscript{6}

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{\textbf{Albery2020PreGlo?}} This network stores
pairwise values of viral community similarity, measured for all hosts
(to maintain consistency with teh phylogenetic diversity measure) across
all viruses; therefore, we consider that it accounts for some overall
similarity in the way hosts deal with viruses, and not only
betacoronaviruses. There is empirical evidence that capacity for
cross-species transmission even between divergent species is generally
high,\textsuperscript{7} especially for
beta-coronaviruses.\textsuperscript{8} To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

\leavevmode\hypertarget{ref-Muylaert2022Present}{}%
\CSLLeftMargin{5. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Ka-WaiHui2006Reasons}{}%
\CSLLeftMargin{6. }
\CSLRightInline{Ka-Wai Hui, E. Reasons for the increase in emerging and
re-emerging viral infectious diseases. \emph{Microbes and Infection}
\textbf{8}, 905--916 (2006).}

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{7. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}

\leavevmode\hypertarget{ref-Latinne2020Origin}{}%
\CSLLeftMargin{8. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}

>>>>

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{5} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{6} or
complexes\textsuperscript{7} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{8} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{9} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{10} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. In their bat reservoirs, coronaviruses evolve through a mix
of host jumps, recombination among disparate lineages, and, to a lesser
degree, co-divergence with their hosts---\textsuperscript{11}a mix of
mechanisms that creates a complex and nonlinear relationship between
host diversity and viral emergence. Working from a recently published
database of bat hosts of betacoronaviruses, we test whether spatial
structure in bat-betacoronavirus coevolution is identifiable at a global
scale. Aiming to explain these patterns, we develop a generalized
framework for applying the GMTC to host-virus interactions, with a
specific emphasis on the potential to create independent coevolutionary
dynamics (and therefore spatial fragmentation in risk) through
heterogeneity. We develop a trivariate risk assessment system that
connects each GMTC mechanism to a quantifiable aspect of host-virus
interactions: (i) viral sharing rates in host communities, representing
the strength of potential interaction between viruses and any one host
(i.e., places where viruses undergo constant host switching may be
coevolutionary coldspots); (ii) the phylogenetic diversity of hosts, as
a proxy for variation in the immunological mechanisms that antagonize
viruses (i.e., the selection mosaic); and (iii) the local uniqueness of
the bat community, representing the potential for viruses to be exposed
to novel host traits (e.g., variation in receptor sequences). Together,
we argue that these can be used to identify and map the evolutionary
drivers that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?}}
Projecting bat and betacoronavirus phylogeny over space
(fig.~\ref{fig:biogeo}), we find support for the idea that bat community
assembly is directly responsible for a global mosaic of viral evolution.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia; (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and Latin America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of multiple
betacoronavirus subgenera, including nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{12}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{11}a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{11,13} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary hosts of the
subgenus \emph{Merbecovirus} (MERS-like viruses);\textsuperscript{12,13}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{14--17} At the coarsest scale, these
heterogeneities are lost, and betacoronavirus biogeography tracks the
deep rifts in bat evolutionary history---but within broad regions, the
component coevolutionary systems may have very different dynamics.

@@

Bats---and the spillover of their viruses---are also sensitive to
anthropogenic factors others than climate change, including
deforestation and other kinds of habitat loss, increased stress, and
greater contact with potential bridge hosts like domesticated
species.\textsuperscript{\textbf{Alves2018GeoVar?},\textbf{Treitler2016EffLoc?},\textbf{Rulli2021LanCha?},\textbf{Mendenhall2014PreBio?}}
This represents a challenge for both conservation strategies and
pandemic prevention,\textsuperscript{\textbf{Amman2011InvRol?}} but
identifying areas at risk, and protecting the health of bats and
ecosystems within those zones, can be a win-win intervention for
both.\textsuperscript{\textbf{Hopkins2021HowIde?},\textbf{Plowright2021LanUse?},\textbf{OHHLEP2022OneHea?}}
As we scale these predictions down in space to finer spatial resolutions
to guide public health actions,\textsuperscript{18} the incorporation of
human activity predictors will become more
importyant.\textsuperscript{19}

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{\textbf{Albery2020PreGlo?}} This network stores
pairwise values of viral community similarity, measured for all hosts
(to maintain consistency with teh phylogenetic diversity measure) across
all viruses; therefore, we consider that it accounts for some overall
similarity in the way hosts deal with viruses, and not only
betacoronaviruses. There is empirical evidence that capacity for
cross-species transmission even between divergent species is generally
high,\textsuperscript{20} especially for
beta-coronaviruses.\textsuperscript{13} To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

\leavevmode\hypertarget{ref-Thompson2005Geographic}{}%
\CSLLeftMargin{5. }
\CSLRightInline{Thompson, J. N. \emph{The Geographic Mosaic of
Coevolution}. (University Of Chicago Press, 2005).}

\leavevmode\hypertarget{ref-Thompson1994Coevolutionary}{}%
\CSLLeftMargin{6. }
\CSLRightInline{Thompson, J. N. \emph{The Coevolutionary Process}.
(University of Chicago Press, 1994).}

\leavevmode\hypertarget{ref-Janzen1980When}{}%
\CSLLeftMargin{7. }
\CSLRightInline{Janzen, D. H. When is it Coevolution? \emph{Evolution}
\textbf{34}, 611--612 (1980).}

\leavevmode\hypertarget{ref-Price2002Macroevolutionary}{}%
\CSLLeftMargin{8. }
\CSLRightInline{Price, P. W. \emph{Macroevolutionary Theory on
Macroecological Patterns}. (Cambridge University Press, 2002).
doi:\href{https://doi.org/10.1017/CBO9780511615030}{10.1017/CBO9780511615030}.}

\leavevmode\hypertarget{ref-Gomulkiewicz2007Dos}{}%
\CSLLeftMargin{9. }
\CSLRightInline{Gomulkiewicz, R. \emph{et al.} Dos and don'ts of testing
the geographic mosaic theory of coevolution. \emph{Heredity}
\textbf{98}, 249--258 (2007).}

\leavevmode\hypertarget{ref-Thrall2007Coevolution}{}%
\CSLLeftMargin{10. }
\CSLRightInline{Thrall, P. H., Hochberg, M. E., Burdon, J. J. \& Bever,
J. D. Coevolution of symbiotic mutualists and parasites in a community
context. \emph{Trends in Ecology \& Evolution} \textbf{22}, 120--126
(2007).}

\leavevmode\hypertarget{ref-Anthony2017Global}{}%
\CSLLeftMargin{11. }
\CSLRightInline{Anthony, S. J. \emph{et al.} Global patterns in
coronavirus diversity. \emph{Virus Evolution} \textbf{3}, (2017).}

\leavevmode\hypertarget{ref-Ruiz-Aravena2022Ecology}{}%
\CSLLeftMargin{12. }
\CSLRightInline{Ruiz-Aravena, M. \emph{et al.} Ecology, evolution and
spillover of coronaviruses from bats. \emph{Nature Reviews Microbiology}
\textbf{20}, 299--314 (2022).}

\leavevmode\hypertarget{ref-Latinne2020Origin}{}%
\CSLLeftMargin{13. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}

\leavevmode\hypertarget{ref-Anthony2013Coronaviruses}{}%
\CSLLeftMargin{14. }
\CSLRightInline{Anthony, S. J. \emph{et al.} Coronaviruses in bats from
Mexico. \emph{The Journal of General Virology} \textbf{94}, 1028--1038
(2013).}

\leavevmode\hypertarget{ref-Goes2013Novel}{}%
\CSLLeftMargin{15. }
\CSLRightInline{Góes, L. G. B. \emph{et al.} Novel Bat Coronaviruses,
Brazil and Mexico. \emph{Emerging Infectious Diseases} \textbf{19},
1711--1713 (2013).}

\leavevmode\hypertarget{ref-Goes2016Genetic}{}%
\CSLLeftMargin{16. }
\CSLRightInline{Góes, L. G. B. \emph{et al.} Genetic diversity of bats
coronaviruses in the Atlantic Forest hotspot biome, Brazil.
\emph{Infection, Genetics and Evolution: Journal of Molecular
Epidemiology and Evolutionary Genetics in Infectious Diseases}
\textbf{44}, 510--513 (2016).}

\leavevmode\hypertarget{ref-Brandao2008Coronavirus}{}%
\CSLLeftMargin{17. }
\CSLRightInline{Brandão, P. E. \emph{et al.} A coronavirus detected in
the vampire bat Desmodus rotundus. \emph{Brazilian Journal of Infectious
Diseases} \textbf{12}, 466--468 (2008).}

\leavevmode\hypertarget{ref-Muylaert2022Present}{}%
\CSLLeftMargin{18. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Ka-WaiHui2006Reasons}{}%
\CSLLeftMargin{19. }
\CSLRightInline{Ka-Wai Hui, E. Reasons for the increase in emerging and
re-emerging viral infectious diseases. \emph{Microbes and Infection}
\textbf{8}, 905--916 (2006).}

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{20. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}
